Integration of national demographic-disturbance relationships and local data can improve caribou population viability projections and inform monitoring decisions

### 2025-02-24 analysis/paper/paper.Rmd
submitted.

institute:
  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, 1125 Colonel By Dr, Ottawa, ON, K1S 5B6, Canada
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/ecological-informatics.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed. There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but no populations inhabiting areas with high anthropogenic disturbance are considered self-sustaining. Demographic projections derived from local population-specific data are uncertain for populations with limited monitoring. We propose a simple Bayesian population model that integrates prior information from a national analysis of demographic-disturbance relationships with available local demographic data to improve population viability projections, and to reduce the risk that a lack of local data will be used as a reason to delay conservation action. The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves, through a term derived from a simple model of the recruitment survey observation process. We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At lower disturbance levels where initial uncertainty is high, each additional year of monitoring provides less new information. The estimated probability of viability indicates whether more information is needed to improve accuracy of population viability projections. 
keywords: |
  woodland caribou (*Rangifer tarandus caribou*); anthropogenic disturbance scenarios; population viability; value of information; monitoring decisions; species at risk; prior information
highlights: |

@@

The number of species threatened with extinction is high and getting higher [@ipbes_global_2019]. Reliable information on the status of populations and likely impacts of management are needed to inform conservation action, but effective monitoring is challenging and costly [@legg_why_2006; @lindenmayer_effective_2018; @wintle_allocating_2010;@dunham_strategic_2023]. Resources to support conservation and recovery are limited [@buxton_balancing_2022; @mccarthy_financial_2012;@gerber_conservation_2016; @wintle_spending_2019], delays in action increase risk [@martin_acting_2012], and some populations have been monitored to extinction [@lindenmayer_counting_2013]. A recent analysis of action plans for Canadian species at risk found that half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022]. Analysis of the implications of monitoring outcomes can help to clarify whether additional monitoring is necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;@gregory_structured_2012].

Boreal caribou are widely and sparsely distributed through the Canadian boreal region across a gradient of anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020] that includes some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022; @ibisch_global_2016;
@venter_sixteen_2016]. Across 58 boreal caribou study areas, both survival and recruitment decrease with the percentage of the area that is disturbed, and no populations inhabiting areas with high anthropogenic disturbance are considered viable in the absence of interventions such as predator control or maternal penning (i.e. self-sustaining) [@eccc_scientific_2011; @johnson_science_2020]. Identifying populations that are declining is important for clarifying where and when conservation action is needed. Disturbance affects caribou demography both through direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022; @eccc_scientific_2011]. Decreasing recruitment with increased disturbance has also been observed over time where historic demographic data exist [@rudolph_demographic_2017; @hervieux_widespread_2013], but many places lack sufficient data to characterize local demographic-disturbance relationships [@johnson_science_2020]. There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020; @neufeld_primary_2021;
@fortin_forest_2017; @gagne_impact_2016], which leads to uncertainty in projections that rely only on national-scale relationships [@stewart_climate-informed_2023; @dyson_existing_2022].

@@

Here, we offer a simple Bayesian population model for boreal caribou (*Rangifer tarandus
caribou*) that integrates prior information about the impacts of landscape change [i.e. demographic-disturbance relationships in @johnson_science_2020] with available demographic data to reduce uncertainty in population viability projections. Note that for the purposes of this analysis we consider a population to be viable if the population growth rate is > 0.99. We also offer a novel method of acknowledging and accounting for errors in composition surveys. Following a value information analysis framework for conservation status assessments [@dunham_strategic_2023], we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023; @dyson_existing_2022] to (i) show how the value of local monitoring varies with landscape condition, and (ii) assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections.

@@

To investigate how the accuracy and precision of predictions vary with both landscape condition and monitoring effort, we analyze simulated data using a Bayesian model with informative priors (section 2.4) and a composition survey bias term (section 2.1). Steps in our analysis procedure include defining anthopogenic disturbance and monitoring scenarios (section 2.2), simulating population dynamics and monitoring (section 2.3) to produce example demographic trajectories for each scenario, and analyzing example demographic trajectories with a Bayesian model that enables integration of (simulated) local demographic data with prior knowledge of demographic-disturbance relationships (section 2.4). Outcomes of interest include expected population growth rate, the probability of correctly assessing population status, the posterior probability of population viability, and the expected value of local information (section 2.5). We refer to the model used to simulate example trajectories as the "national" model because parameters are from a national analysis of demographic-disturbance relationships [@johnson_science_2020]. We distinguish this national model from the Bayesian model that is used to analyze simulated example demographic trajectories. Where appropriate the national model and the Bayesian model share the same structure, and we have selected notation to highlight the similarities. Bayesian model estimates are distinguished from observed values, estimated expected values, true values, and true expected values using $\hat{\text{ }}$ $\tilde{\text{ }}$ $\dot{\text{ }}$ and $\bar{\text{ }}$ symbols respectively; for example, $S_t$ is the Bayesian estimate of survival in year $t$, $\hat{S}_t$ is observed survival, $\tilde{S}_t$ is the Bayesian estimate of expected survival (without stochastic interannual variation), $\dot{S}_t$ is true survival, and $\bar{S_t}$ is true expected survival (without stochastic interannual variation). For reference, all variables and parameters are defined in supplemental Table A1.

@@

The most widely available source of boreal caribou demographic
information is from cows fitted with telemetry collars [e.g.
@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
@decesare_estimating_2012], though genetic methods are playing an
increasingly important role in some caribou monitoring programs [e.g. @flasko_fecal_2017;
@hettinga_estimating_2012; @mcfarlane_genetic_2018;
@moeller_integrating_2021; @jones_novel_2023;
@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
are typically captured via net-gun from aircraft in late winter, and
affixed with collars that send a mortality signal when no longer
active [see @nbckc_nbckc_2021], thus allowing estimation of adult
survival [e.g. @dalgarno_bboutools_2024; @mcloughlin_declines_2003;
@hervieux_widespread_2013; @eacker_web-based_2019; @lamb_effectiveness_2024]. Recruitment is estimated from
aerial surveys of the composition of groups of animals that include
collared animals [@nbckc_nbckc_2022]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that different demographic groups are equally detectable
[@ellington_divergent_2020]. Errors that can bias composition survey
results include misidentifying young bulls as cows or vice versa , and
failing to detect calves [@ellington_divergent_2020;
@decesare_estimating_2012]. Survival and recruitment survey data can be combined in a simple
recruitment-mortality population model [@hatter_moose_1991;@hatter_revisiting_2020] that serves
as the basis for many herd viability assessments
[@decesare_estimating_2012; @serrouya_comparing_2017;
@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020; @dalgarno_bboutools_2024].

@@

We focus on exploring a set of scenarios in which buffered anthropogenic
disturbance, measured as the percentage of area within 500 m of some
type of anthropogenic disturbance [@johnson_science_2020;
@eccc_scientific_2011], increases by 1% per year (Figure
\@ref(fig:figDisturbance)). The duration of monitoring prior to the
projection period varies among scenarios from 1 to 24 years. The
projection period begins in 2024, and we consider how uncertainty
changes as the projection period increases from 5 to 20 years. The
amount of anthropogenic disturbance at the beginning of the projection
period varies from low (0%) through medium-low (20%) and medium-high
(40%) to high (60%). These simple hypothetical scenarios span a large
range to show how the need for local monitoring to reduce uncertainty
depends on landscape state, not to represent the reality of landscape
change. Scenarios in which anthropogenic disturbance remains the same or
decreases over time are included in Supplement D.

@@

In both models recruitment probability is a function of anthropogenic
disturbance and fire with a log link and Beta distributed interannual variation [@johnson_science_2020;
@stewart_climate-informed_2023; @dyson_existing_2022]. The coefficient of variation $\nu_R$ is constant, while the mean $\tilde{R_t}$ and standard deviation $\sigma_{R,t}$ vary with disturbance:
\begin{equation}
log(\tilde{R_t})=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t
  (\#eq:Rtilde)
\end{equation}

@@

As done by Johnson et al. [-@johnson_science_2020], we model survival
probability as an adjusted function of buffered anthropogenic
disturbance $A_t$ with a log link and Beta distributed interannual variation with coefficient of variation $\nu_S$, mean $\tilde{S_t}$ and standard deviation $\sigma_{S,t}$ [@stewart_climate-informed_2023;
@dyson_existing_2022]: 
\begin{equation}
\tilde{S_t}=(46 e^{\beta^S_0+\beta^S_a A_t}-0.5)/45
  (\#eq:Stilde)
\end{equation}

@@

All regression coefficients are assumed to be Gaussian distributed, with
prior means from Johnson et al's (2020) regression models
(Table \@ref(tab:priors)). We calibrated the standard deviations of the
priors so that the the 95% prior predictive intervals for survival and
recruitment from the Bayesian model match the 95% prediction intervals
from Johnson et al's regression model (Supplement C, Figure
\@ref(fig:examplePredictions)). To show the value of informative priors, we also include a weakly informative scenario with prior means of the intercept parameters from Dalgarno et al. [-@dalgarno_bboutools_2024] ($\beta^R_0=-0.3689, \beta^S_0=-0.2126776$), prior means of the slope parameters set to zero, and standard deviations for all parameters (Table \@ref(tab:priors)) multiplied by ten.

@@

Analysis of simulated example demographic trajectories confirms that our
Bayesian model can effectively combine local demographic data with
priors from a national demographic-disturbance relationships to reduce
uncertainty in boreal caribou demographic projections (Figures
\@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). When local data are limited (e.g. 4 years, 15 collars per year) the model with informative priors (i.e. knowledge of national relationships) yields predictions that are consistent with knowledge of observed outcomes across the country, and the model with weakly informative priors does not (Figure \@ref(fig:examplePredictionsLowEffort)). When there is interannual variation increasing the duration of monitoring improves predictions of population growth rate substantially, and increasing the number of collars each year is less effective (Figure \@ref(fig:figDiffs)). Insufficiently informative priors can yield predictions that are both imprecise and inaccurate even in a case where there are 16 years of local monitoring data (Figure \@ref(fig:examplePredictions)), highlighting the benefits of integrating national relationships with these local data. Increasing the forecast horizon [@petchey_ecological_2015] does not substantially decrease the accuracy of expected population growth projections (compare 0, 5 and 20 year projections in
Figure \@ref(fig:figDiffs)) in these scenarios.

@@

Interventions such as wolf culls, maternal penning, and restoration of
linear features are intended to improve survival and
reproduction of populations that are not otherwise self-sustaining
[@mcnay_demographic_2022; @serrouya_predicting_2020], and there is a
need to evaluate and compare effects of boreal caribou conservation
actions with careful monitoring. Our focus in this paper is on assessing
and projecting viability of populations in the absence of
these interventions, to clarify when additional information is needed to
reduce uncertainty about whether there is a need for action. The model we
propose does not include effects of management interventions, and
cannot be used to project outcomes for managed populations. Our intention is to reduce
uncertainty that can be used to delay conservation action; there is
nothing in our arguments, models or results that can or should be used
to discourage monitoring of the effectiveness of conservation
actions. Even when no additional monitoring is required for assessing
viability, there are a variety of reasons to continue
monitoring highly disturbed boreal caribou populations. The point we wish to emphasize is
that action is also required, and lack of monitoring should not be used
as a reason to delay conservation measures for highly disturbed
populations.

Variation in anthropogenic disturbance across the Canadian boreal region
is partially correlated with variation in climate and primary
productivity [@neilson_direct_2022] because forestry occurs in higher
productivity areas, and petroleum extraction activities are concentrated
in Western Canada. Effects of disturbance on vegetation and foodweb
dynamics vary with primary productivity [@neufeld_primary_2021], and
some have expressed concern about confounding of demographic-disturbance
relationships [@wilson_causal_2021]. One option for addressing these
concerns would be to develop regional demographic-disturbance
relationships for populations that are ecologically similar to one
another (e.g. caribou inhabiting forests that are sufficiently
productive for industrial forestry in Eastern Canada)
[@eccc_scientific_2024], and to use the priors from regional
demographic-disturbance relationships to inform demographic projections
within the region. However, while acknowledging and allowing for
regional variation, we also need to guard against risks that uncertainty
will be inflated and used as an excuse to delay action, a tactic that
has been used in this and other decision contexts
[@bissonnette_controverses_2020; @boan_climate_2018]. In order to be
informative and credible, regional demographic-disturbance analyses must
include a sufficient number of observations spanning a range from high
to low anthropogenic disturbance. Priors that lead to projections of
outcomes for caribou that fall outside the observed range of variation
across the country, characterized here (Figure 2) by 95% prediction
intervals from a national analysis (Johnson et al. 2020), are
implausible. In other words, if we have seen no examples of
self-sustaining caribou populations at a particularly high level of
disturbance, it is unreasonable to project or expect this outcome.

@@

Aspatial demographic models generally assume that observed animals are a
random and representative sample of a fully mixed population in which
survival and recruitment rates apply equally to all animals. However,
metrics derived from fecal pellet analysis such as reproductive success
[@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;
@mcfarlane_identification_2022], age structure and pregnancy rates
[@flasko_fecal_2017; @mcfarlane_identification_2022], and
mother-offspring co-occurence [@jones_novel_2023] suggest high variation
in fitness among individuals [@mcfarlane_genetic_2018] and spatial
structure within populations [@jones_novel_2023;
@mcfarlane_spatial_2021; @mcfarlane_identification_2022]. Analyses of
movement data also show spatial variation in risks and benefits for
caribou and variation in habitat preferences and space use among
individuals [e.g. @avgar_spaceuse_2015; @decesare_linking_2014]. It is possible to integrate complexities such as variation in demographic rates
among individuals [@armstrong_incorporating_2021], population size
information from aerial or fecal DNA surveys, and variation in demographic rates among age classes [@moeller_integrating_2021; @mcnay_demographic_2022] into Bayesian population models. However, in many contexts there are not enough to data to estimate more complex model parameters, and simple models remain relevant.

Our proof-of-concept analysis shows the value of integrating national demographic-disturbance relationships with local data, accounting for composition survey errors, and assessing the value of local monitoring
information. We applied the framework proposed by Dunham et al. (2023) to show that
no additional local monitoring information is needed to predict whether
boreal caribou populations in highly disturbed landscapes are viable in the absence of interventions such as predator control or maternal penning; self-sustainability is unlikely. There is more variation in caribou demographic rates
among study areas with low to moderate disturbance, and thus more need
for local data to inform demographic projections. In these places, our
analysis suggests that there may be diminishing returns from
increasing monitoring effort, but more information on monitoring costs,
objectives, and existing local data would be required to assess the
adequacy and cost-effectiveness of monitoring scenarios in particular
circumstances. Monitoring costs vary substantially with technologies
used, the remoteness of herds, and the difficulty of collaring animals,
which in turn depends on the number of currently collared animals, the
population size, and vegetation across the landscape
[@nbckc_boreal_2021]. It is important to recognize that projecting
population viability is only one possible objective for monitoring
[@nbckc_boreal_2019], and that the impacts of monitoring on threatened
animals should be considered [@nbckc_boreal_2019]. Since monitoring
costs, options and goals are context specific, we believe the most
effective way forward will be to use this proof-of-concept analysis, existing local data, and the more fully developed *bboutools* R package [@dalgarno_bboutools_2024] as a starting point for working with interested decision makers to develop monitoring decision support tools that meet real local needs, recognizing that those needs will vary [@nbckc_nbckc_2021;@nbckc_nbckc_2022].

@@

# Data statement
Code required to create the data, run analyses and reproduce figures in this paper is available at https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper, and in the caribouMetrics R package available on GitHub at https://github.com/LandSciTech/caribouMetrics/tree/EI-paper-submission.

>>>>

institute:
  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, 1125 Colonel By Dr, Ottawa, ON, K1S 5B6, Canada
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/ecological-informatics.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou study areas in Canada, survival and recruitment decrease with the percentage of the study area that is disturbed. There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but no populations inhabiting areas with high anthropogenic disturbance are considered viable. Demographic projections derived from local population-specific data are uncertain for populations with limited monitoring. We propose a simple Bayesian population model that integrates prior information from a national analysis of demographic-disturbance relationships with available local demographic data to improve population viability projections, and to reduce the risk that a lack of local data will be used as a reason to delay conservation action. The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves, through a term derived from a simple model of the recruitment survey observation process. We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At lower disturbance levels where initial uncertainty is high local data improve accuracy but each additional year of monitoring provides less new information. The estimated probability of viability indicates whether more information is needed to improve accuracy of population viability projections. 
keywords: |
  woodland caribou (*Rangifer tarandus caribou*); anthropogenic disturbance scenarios; population viability; value of information; monitoring decisions; species at risk; prior information
highlights: |

@@

The number of species threatened with extinction is high and getting higher [@ipbes_global_2019]. Reliable information on the status of populations and likely impacts of management are needed to inform conservation action, but effective monitoring is challenging and costly [@legg_why_2006; @lindenmayer_effective_2018; @wintle_allocating_2010;@dunham_strategic_2023]. Resources to support conservation and recovery are limited [@buxton_balancing_2022; @mccarthy_financial_2012;@gerber_conservation_2016; @wintle_spending_2019], delays in action increase risk [@martin_acting_2012], and some populations have been monitored to extinction [@lindenmayer_counting_2013]. A recent analysis of action plans for Canadian species at risk found that half of recovery actions are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022]. Analysis of the implications of monitoring outcomes can help to clarify whether additional monitoring is necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;@gregory_structured_2012].

Boreal caribou are widely and sparsely distributed through the Canadian boreal region across a gradient of anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020] that includes some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022; @ibisch_global_2016;
@venter_sixteen_2016]. Across 58 boreal caribou study areas, both survival and recruitment decrease with the percentage of the area that is disturbed, and no populations inhabiting areas with high anthropogenic disturbance are considered viable, defined as a population  experiencing stable or positive growth (≥0.99) in the absence of interventions such as predator control or maternal penning [@eccc_scientific_2011; @johnson_science_2020]. We acknowledge that other considerations such as population size are important, but our focus in this analysis is population growth rate. Identifying populations that are declining is important for clarifying where and when conservation action is needed. Disturbance affects caribou demography both through direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022; @eccc_scientific_2011]. Decreasing recruitment with increased disturbance has also been observed over time where historic demographic data exist [@rudolph_demographic_2017; @hervieux_widespread_2013], but many places lack sufficient data to characterize local demographic-disturbance relationships [@johnson_science_2020]. There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020; @neufeld_primary_2021;
@fortin_forest_2017; @gagne_impact_2016], which leads to uncertainty in projections that rely only on national-scale relationships [@stewart_climate-informed_2023; @dyson_existing_2022].

@@

Here, we offer a simple Bayesian population model for boreal caribou (*Rangifer tarandus
caribou*) that integrates prior information about the impacts of landscape change [i.e. demographic-disturbance relationships in @johnson_science_2020] with available demographic data to reduce uncertainty in population viability projections. We also offer a novel method of acknowledging and accounting for errors in composition surveys. Following a value information analysis framework for conservation status assessments [@dunham_strategic_2023], we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023; @dyson_existing_2022] to (i) show how the value of local monitoring varies with landscape condition, and (ii) assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections.

@@

To investigate how the accuracy and precision of predictions vary with both landscape condition and monitoring effort, we analyze simulated data using a Bayesian model with informative priors (section 2.4) and a composition survey bias term (section 2.1). Steps in our analysis procedure include defining anthropogenic disturbance and monitoring scenarios (section 2.2), simulating population dynamics and monitoring (section 2.3) to produce example demographic trajectories for each scenario, and analyzing example demographic trajectories with a Bayesian model that enables integration of (simulated) local demographic data with prior knowledge of demographic-disturbance relationships (section 2.4). Outcomes of interest include expected population growth rate, the probability of correctly assessing population status, the posterior probability of population viability, and the expected value of local information (section 2.5). We refer to the model used to simulate example trajectories as the "national" model because parameters are from a national analysis of demographic-disturbance relationships [@johnson_science_2020]. We distinguish this national model from the Bayesian model that is used to analyze simulated example demographic trajectories. Where appropriate the national model and the Bayesian model share the same structure, and we have selected notation to highlight the similarities. Bayesian model estimates are distinguished from observed values, estimated expected values, true values, and true expected values using $\hat{\text{ }}$ $\tilde{\text{ }}$ $\dot{\text{ }}$ and $\bar{\text{ }}$ symbols respectively; for example, $S_t$ is the Bayesian estimate of survival in year $t$, $\hat{S}_t$ is observed survival, $\tilde{S}_t$ is the Bayesian estimate of expected survival (without stochastic interannual variation), $\dot{S}_t$ is true survival, and $\bar{S_t}$ is true expected survival (without stochastic interannual variation). For reference, all variables and parameters are defined in supplemental Table A1.

@@

The most widely available source of boreal caribou demographic
information is from cows fitted with telemetry collars [e.g.
@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
@decesare_estimating_2012], though genetic methods are playing an
increasingly important role in some caribou monitoring programs [e.g. @flasko_fecal_2017;
@hettinga_estimating_2012; @mcfarlane_genetic_2018;
@moeller_integrating_2021; @jones_novel_2023;
@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
are typically captured via net-gun from aircraft in late winter, and
affixed with collars that send a mortality signal when no longer
active [see @nbckc_nbckc_2021], thus allowing estimation of adult
survival [e.g. @dalgarno_bboutools_2024; @mcloughlin_declines_2003;
@hervieux_widespread_2013; @eacker_web-based_2019; @lamb_effectiveness_2024]. Recruitment is estimated from
aerial surveys of the composition of groups of animals that include
collared animals [@nbckc_nbckc_2022]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that different demographic groups are equally detectable
[@ellington_divergent_2020]. Errors that can bias composition survey
results include misidentifying young bulls as cows or vice versa, and
failing to detect calves [@ellington_divergent_2020;
@decesare_estimating_2012]. Survival and recruitment survey data can be combined in a simple
recruitment-mortality population model [@hatter_moose_1991;@hatter_revisiting_2020] that serves
as the basis for many herd viability assessments
[@decesare_estimating_2012; @serrouya_comparing_2017;
@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020; @dalgarno_bboutools_2024].

@@

We focus on exploring a set of scenarios in which anthropogenic
disturbance, measured as the percentage of area within 500 m of some
type of anthropogenic disturbance [@johnson_science_2020;
@eccc_scientific_2011], increases by 1% per year (Figure
\@ref(fig:figDisturbance)). The duration of monitoring prior to the
projection period varies among scenarios from 1 to 24 years. The
projection period begins in 2024, and we consider how uncertainty
changes as the projection period increases from 5 to 20 years. The
amount of anthropogenic disturbance at the beginning of the projection
period varies from low (0%) through medium-low (20%) and medium-high
(40%) to high (60%). These simple hypothetical scenarios span a large
range to show how the need for local monitoring to reduce uncertainty
depends on landscape state, not to represent the reality of landscape
change. Scenarios in which anthropogenic disturbance remains the same or
decreases over time are included in Supplement D.

@@

In both models, recruitment probability is a function of anthropogenic
disturbance and fire with a log link and Beta distributed interannual variation [@johnson_science_2020;
@stewart_climate-informed_2023; @dyson_existing_2022]. The coefficient of variation $\nu_R$ is constant, while the mean $\tilde{R_t}$ and standard deviation $\sigma_{R,t}$ vary with disturbance:
\begin{equation}
log(\tilde{R_t})=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t
  (\#eq:Rtilde)
\end{equation}

@@

As done by Johnson et al. [-@johnson_science_2020], we model survival
probability as an adjusted function of anthropogenic
disturbance $A_t$ with a log link and Beta distributed interannual variation with coefficient of variation $\nu_S$, mean $\tilde{S_t}$ and standard deviation $\sigma_{S,t}$ [@stewart_climate-informed_2023;
@dyson_existing_2022]: 
\begin{equation}
\tilde{S_t}=(46 e^{\beta^S_0+\beta^S_a A_t}-0.5)/45
  (\#eq:Stilde)
\end{equation}

@@

All regression coefficients are assumed to be Gaussian distributed, with
prior means from Johnson et al's (2020) regression models
(Table \@ref(tab:priors)). We calibrated the standard deviations of the
priors so that the the 95% prior predictive intervals for survival and
recruitment from the Bayesian model match the 95% prediction intervals
from Johnson et al's (2020) regression model (Supplement C, Figure
\@ref(fig:examplePredictions)). To show the value of informative priors, we also include a weakly informative scenario with prior means of the intercept parameters from Dalgarno et al. [-@dalgarno_bboutools_2024] ($\beta^R_0=-0.3689, \beta^S_0=-0.2126776$), prior means of the slope parameters set to zero, and standard deviations for all parameters (Table \@ref(tab:priors)) multiplied by ten.

@@

Analysis of simulated example demographic trajectories confirms that our
Bayesian model can effectively combine local demographic data with
priors from a national demographic-disturbance relationships to reduce
uncertainty in boreal caribou demographic projections (Figures
\@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). When local data are limited (e.g. 4 years, 15 collars per year) the model with informative priors (i.e. knowledge of national relationships) yields predictions that are consistent with knowledge of observed outcomes across the country, and the model with weakly informative priors does not (Figure \@ref(fig:examplePredictionsLowEffort)). When there is interannual variation, increasing the duration of monitoring improves predictions of population growth rate substantially, and increasing the number of collars each year is less effective (Figure \@ref(fig:figDiffs)). Insufficiently informative priors can yield predictions that are both imprecise and inaccurate even in a case where there are 16 years of local monitoring data (Figure \@ref(fig:examplePredictions)), highlighting the benefits of integrating national relationships with these local data. Increasing the forecast horizon [@petchey_ecological_2015] does not substantially decrease the accuracy of expected population growth projections (compare 0, 5 and 20 year projections in
Figure \@ref(fig:figDiffs)) in these scenarios.

@@

Interventions such as wolf culls, maternal penning, and restoration of
linear features are intended to improve survival and
reproduction of populations that are not otherwise viable
[@mcnay_demographic_2022; @serrouya_predicting_2020]. Our focus in this paper is on assessing
and projecting viability of populations in the absence of
these interventions, to clarify when additional information is needed to
reduce uncertainty about whether there is a need for action. The model we
propose does not include effects of management interventions, and
cannot be used to project outcomes for managed populations. There is
nothing in our arguments, models or results that can or should be used
to discourage monitoring of the effectiveness of conservation
actions. The point we wish to emphasize is
that action is also required, and lack of monitoring should not be used
as a reason to delay conservation measures for highly disturbed
populations.

Variation in anthropogenic disturbance across the Canadian boreal region
is partially correlated with variation in climate and primary
productivity [@neilson_direct_2022] because forestry occurs in higher
productivity areas, and petroleum extraction activities are concentrated
in Western Canada. Effects of disturbance on vegetation and foodweb
dynamics vary with primary productivity [@neufeld_primary_2021], and
some have expressed concern about confounding of demographic-disturbance
relationships [@wilson_causal_2021]. One option for addressing these
concerns would be to develop regional demographic-disturbance
relationships for populations that are ecologically similar to one
another (e.g. caribou inhabiting forests that are sufficiently
productive for industrial forestry in Eastern Canada)
[@eccc_scientific_2024], and to use the priors from regional
demographic-disturbance relationships to inform demographic projections
within the region. However, while acknowledging and allowing for
regional variation, we also need to guard against risks that uncertainty
will be inflated and used as an excuse to delay action, a tactic that
has been used in this and other decision contexts
[@bissonnette_controverses_2020; @boan_climate_2018]. In order to be
informative and credible, regional demographic-disturbance analyses must
include a sufficient number of observations spanning a range from high
to low anthropogenic disturbance. Priors that lead to projections of
outcomes for caribou that fall outside the observed range of variation
across the country, characterized here (Figure 2) by 95% prediction
intervals from a national analysis (Johnson et al. 2020), are
implausible. In other words, if we have seen no examples of
viable caribou populations at a particularly high level of
disturbance, it is unreasonable to project or expect this outcome.

@@

Aspatial demographic models generally assume that observed animals are a
random and representative sample of a fully mixed population in which
survival and recruitment rates apply equally to all animals. However,
metrics derived from fecal pellet analysis such as reproductive success
[@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;
@mcfarlane_identification_2022], age structure and pregnancy rates
[@flasko_fecal_2017; @mcfarlane_identification_2022], and
mother-offspring co-occurence [@jones_novel_2023] suggest high variation
in fitness among individuals [@mcfarlane_genetic_2018] and spatial
structure within populations [@jones_novel_2023;
@mcfarlane_spatial_2021; @mcfarlane_identification_2022]. Analyses of
movement data also show spatial variation in risks and benefits for
caribou and variation in habitat preferences and space use among
individuals [e.g. @avgar_spaceuse_2015; @decesare_linking_2014]. It is possible to integrate complexities such as variation in demographic rates
among individuals [@armstrong_incorporating_2021], population size
information from aerial or fecal DNA surveys, and variation in demographic rates among age classes [@moeller_integrating_2021; @mcnay_demographic_2022] into Bayesian population models. However, in many contexts there are not enough data to estimate more complex model parameters, and simple models remain relevant.

Our proof-of-concept analysis shows the value of integrating national demographic-disturbance relationships with local data, accounting for composition survey errors, and assessing the value of local monitoring
information. We applied the framework proposed by Dunham et al. (2023) to show that
no additional local monitoring information is needed to predict whether
boreal caribou populations in highly disturbed landscapes are viable in the absence of interventions such as predator control or maternal penning; viability is unlikely. There is more variation in caribou demographic rates
among study areas with low to moderate disturbance, and thus more need
for local data to inform demographic projections. In these places, our
analysis suggests that there may be diminishing returns from
increasing monitoring effort, but more information on monitoring costs,
objectives, and existing local data would be required to assess the
adequacy and cost-effectiveness of monitoring scenarios in particular
circumstances.

Monitoring costs vary substantially with technologies
used, the remoteness of herds, and the difficulty of collaring animals,
which in turn depends on the number of currently collared animals, the
population size, and vegetation across the landscape
[@nbckc_boreal_2021]. It is important to recognize that projecting
population viability is only one possible objective for monitoring
[@nbckc_boreal_2019], and that the impacts of monitoring on threatened
animals should be considered [@nbckc_boreal_2019]. Since monitoring
costs, options and goals are context specific, we believe the most
effective way forward will be to use this proof-of-concept analysis, existing local data, and the more fully developed *bboutools* R package [@dalgarno_bboutools_2024] as a starting point for working with interested decision makers to develop monitoring decision support tools that meet real local needs, recognizing that those needs will vary [@nbckc_nbckc_2021;@nbckc_nbckc_2022].

@@

# Data statement
Code required to create the data, run analyses and reproduce figures in this paper is available at https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper, and in the caribouMetrics R package available on GitHub at https://github.com/LandSciTech/caribouMetrics/tree/EI-paper-submission. Results are available at https://osf.io/ryn59/.

# Acknowledgements
We thank two anonymous reviewers for their insightful comments and helpful suggestions. This work was informed by a variety of discussions with more people than we can name, but here are a few. Members of the Monitoring Working Group of the National Boreal Caribou Knowledge Consortium and the Scientific Assessment of Federal and Provincial Frameworks for the Conservation of Boreal Caribou in Ontario helped us understand context. Tal Avgar and Micheline Manseau encouraged us to think about composition survey errors. Comments from Joseph Bennett and Daniel Eacker on early versions of this work motivated a variety of improvements. Note that our acknowledgement of these contributions does not imply endorsement; we also acknowledge disagreement on some points, and that all remaining errors are our own.Funding for this work was provided by the Wildlife and Landscape Science Directorate of Environment and Climate Change Canada.
